Retrofitting Effect Handlers onto OCaml

### 2020-11-20 retro-concurrency.tex
squeeze in numbers of major collections for the micro benchmarks with no perform

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. These
programs do not |perform| effects. We also include a version of the same
benchmarks implemented using concurrency monad~\cite{Claessen99} (|monad|) as a
proxy for CPS versions. Recall that the OCaml compiler does not use CPS in its IR.
In the |monad| version, we use a |fork| to invoke the non-tail call and use an
|MVar| to collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The concurrency
monad suffers due to the heap allocation of continuation frames (which need to
be garbage collected), whereas effect handlers benefit from stack allocation of
the frames. Our concurrency monad (and other monadic concurrency libraries such
as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides -- due to
the lack of a stack, exceptions, backtraces, and DWARF unwinding are no longer
useful.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we also implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, whereas the |monad| version was 8.69$\times$ slower than |cps|.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library that spawns a thread per
request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
style and does not have the notion of a thread per request. For comparison, we
also include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.

>>>>

Next, we aim to quantify the overhead of setting up and tearing down effect
handlers compared to a non-tail function call. To this end, we surround the
non-tail calls in the recursive micro benchmarks with an effect handler. These
programs do not |perform| effects. We also implemented the same benchmarks using
a concurrency monad~\cite{Claessen99} (|monad|) as a proxy for CPS versions.
Recall that the OCaml compiler does not use CPS in its IR. In the |monad|
version, we use a |fork| to invoke the non-tail call and use an |MVar| to
collect its result.

The results are presented in Table~\ref{tab:micro_noperform}. They show that
using effect handlers (concurrency monad) is 10.02$\times$ (67.09$\times$) more
expensive than the idiomatic implementation using non-tail calls. The
concurrency monad suffers due to the heap allocation of continuation frames
(which need to be garbage collected), whereas effect handlers benefit from stack
allocation of the frames. For example the number of major collections for the
\texttt{ack} benchmark is 0 for stock OCaml, 1 for \texttt{MC} and 112 for
\texttt{monad}. Our concurrency monad (and other monadic concurrency libraries
such as Lwt~\cite{lwt} and Async~\cite{async}) also have other downsides --
exceptions, backtraces, and DWARF unwinding are no longer useful due to the lack
of a stack.

@@

\textbf{\textit{Generators.}} Generators allow data structures to be traversed on
demand. Many languages including JavaScript and Python provide generators as a
built-in primitive. Using effect handlers (|MC|), given any data structure
(|'a t|) and its iterator (|val iter: 'a t -> ('a -> unit) -> unit|), we can
derive its generator function (|val next : unit -> 'a option|). The code is
available in the supplementary material. We evaluate the performance of
traversing a complete binary tree of depth 25 using this generator. This
involves $2^{26}$ stack switches in total. For comparison, we implemented
a hand-written, selective CPSed~\cite{Nielson01},
defunctionalised~\cite{Danvy01} version (|cps|) and a concurrency monad
(|monad|) version of the generator for the tree. Both |cps| and |monad|
versions are specialised to the binary tree with the usual caveats of
not using the stack for function calls. We observed that the |cps| version was
the fastest, thanks to specialisation and hand optimisation. |MC| version was
only 2.76$\times$ slower than |cps| while being a generic
solution, and the |monad| version was 8.69$\times$ slower than |cps|.

@@

\textbf{\textit{Web server.}} As one of our primary goals of adding effect
handlers to OCaml is to support direct-style concurrency, we implement a
full-fledged HTTP/1.1 web server using effect handlers (|MC|). We use effect
handlers to implement a user-level threading library that spawns a thread per
request. We use \texttt{httpaf}~\cite{httpaf} for HTTP handling, and
\texttt{libev}~\cite{libev} for the eventloop. We compare our implementation
against an Lwt version (|lwt|) which also uses \texttt{httpaf} and
\texttt{libev}. Unlike using effect handlers, the Lwt version is written in monadic
style and does not have the notion of a thread per request. For comparison, we
include a Go 1.13 version (|go|) that uses the
\texttt{net/http}~\cite{nethttp} package. As both the OCaml versions are single
threaded, the Go benchmark is run with \texttt{GOMAXPROCS=1}.
